Constrained, yet diverse: Phylogenetic and morphological filtering in epiphytic ferns

### 2019-08-11 ms/manuscript.Rmd
Update MS
- Add @Tsutsumi2013 for gameto evo
- Add description of CWMs
- Add mention of tree in MPD/MNTD
- Cut text in conclusion, discussion, and methods (null model)

Despite the ecological opportunities afforded by life in the canopy, including escape from disturbance [@Page2002] and abundant light for photosynthesis [@Benzing2000], epiphytes must contend with greater stress in the form of increased evaporation and extreme oscillations in temperature relative to terrestrial plants [@Zotz2001; @Lowman2012]. Associated with this change, epiphytic ferns have evolved various morphological adaptations that minimize water loss from their leaves, including shorter stipes [@WatkinsJr.2010], thicker leaves [@Kluge2007; @WatkinsJr.2007b], and protection with scales or hairs [@WatkinsJr.2006b]. Some lineages have specialized adaptations to epiphytic growth, such as humus-collection (e.g., *Asplenium*; @Benzing1990, *Drynaria*; @Janssen2005), vegetative desiccation tolerance (DT) (e.g., Hymenophyllaceae; @Shreve1911, *Pleopeltis*; @Pessin1925), CAM photosynthesis (e.g., *Pyrrosia*; @Wong1976), and symbiotic relationships with ants (e.g., *Antrophyum*; @WatkinsJr.2008, *Lecanopteris*; @Gay1993a). Many of these same adaptations are found in epiphytic seed plants [@Benzing1990]. However, the evolution of epiphytism in ferns was more complex given that ferns rely on a free-living gametophyte. The fern gametophyte is anatomically simple, and differs from the sporophyte in anatomy, morphology, and physiology. Such differences create a system of 'dual ecology' where selective pressures acting on one stage are different than those acting on another. For example, DT is prevalent in gametophytes of multiple lineages of epiphytic ferns, but is rare in sporophytes [@WatkinsJr.2007c]. Despite evidence for adaptations to epiphytism in multiple lineages in both sporophytes and gametophytes, no studies to our knowledge have analyzed the evolution of such traits across the fern phylogeny including both life stages and representatives of all five Cretaceous epiphytic radiations *sensu* [@Schuettpelz2009].

@@

We used distance-based methods to analyze community phylogenetic and functional diversity. Mean phylogenetic distance (MPD) and mean nearest taxon distance (MNTD) describe the mean distance between all species in a community (MPD) or the mean distance between the closest pairs of species in a community (MNTD) [@Webb2000; Webb2002]. Although these metrics were originally formulated for phylogenetic distances, they can be applied in the same way to trait-based distances, enabling a direct comparison between the two [@Sessa2018]. We log-transformed traits as needed to account for skew, scaled each trait by its standard deviation, excluded frond width and length (both frond size traits highly correlated with stipe length), then calculated the distance matrix using Gower's dissimilarity coeffecient [@Gower1971] with the \function{gowdis} function in the R pacakge \package{FD} [@Laliberte2011], with equal weighting for sporophyte and gametophyte traits. To test for the effect of different trait selection schemes, we compared the results of running the same analyses for various sets of traits with and without log-transformation (Supporting Information).

Raw values of MPD and MNTD are statistically related to species richness [@Swenson2014]; to obtain metrics that are uncorrelated with species richness, we calculated their standard effect sizes (SES) by comparing observed values to a distribution of \num{999} randomized communities. We constructed null communities using the "independent swap" algorithm of @Gotelli2000 that maintains species frequency and abundance while randomizing species identities. We included all species of ferns from Moorea and the neighboring island of Tahiti with genetic data available (`r Ntip(phy)` \si{\species} total; [@Nitta2017]) for phylogenetic null communities, and all species with trait data available (`r fern_traits %>%pull(species) %>% n_distinct`) for functional null communities. We used the \function{ses.mpd} and \function{ses.mntd} functions in the R package \package{picante} [@Kembel2010] to calculate phylogenetic and functional diversity metrics. Hereafter, "MPD~phy~" and "MNTD~phy~", and "MPD~func~" and "MNTD~func~", refer to the standard effect sizes of MPD and MNTD as measured with phylogenetic and functional trait distances, respectively.

We tested the effects of climate and growth habit on diversity metrics using a full-subsets multiple regression approach with general additive models (GAMs) as implemented in the \package{FSSgamm} R package [@Fisher2018]. For each diversity metric, we built a set of models (Gaussian family, with site as a random effect) taking into account all combinations of explanatory variables and their pairwise interactions, and ranked these by AIC~c~ to find the models that best explain the data. Our dataset included `r nrow(div_metrics_all) - nrow(climate_select)` communities that lacked climate data and could not be included in full-subsets multiple regression analysis. Therefore, we also constructed linear models for each diversity metric related to elevation including all sites using the same model selection procedure as in the climate analysis. We checked for spatial autocorrelation in the model residuals with Moran's *I* [@Moran1950] as implemented in the \package{spdep} R package [@Bivand2013].

@@

Results for functional community diversity contradicted our hypotheses; epiphytic communities were more functionally diverse than terrestrial as measured by both MPD and MNTD (\fig \@ref(fig:comm-div)). However, this result is highly dependent on the traits used to measure functional diversity; when the same analyses were done using only size-related traits (scaled, untransformed stipe length, frond length and width, and rhizome diameter), the opposite result was observed (\fig S1). Futhermore, terrestrial species show a much wider range of values for size traits (\fig \@ref(fig:cwm-div)), and neither functional group appears to be restricted to a smaller area in trait space (\fig \@ref(fig:pca)). This suggests that epiphytic ferns are subject to extreme environmental filtering for size-related traits, while simultaneously diversifying along other trait axes. This could be maintained by niche partitioning within epiphytic habitats. Epiphytic habitats are highly heterogeneous, and microclimatic conditions vary with height along the host and distance from the trunk [@Hietz1995]. The vertical stratification of epiphytic species into different zones along height gradients in host trees is well documented [@Zotz2007]. In a study comparing terrestrial and epiphytic ferns along an elevational gradient from \SIrange{30}{3000}{\meter} in Costa Rica, @WatkinsJr.2006a found that epiphytes tended to sort into low-trunk and canopy zones, with only 18\% found in both. The goal of our study was to quantify differences between epiphytic and terrestrial fern communities; future research should focus on bringing comparative phylogenetic and community ecological methods to bear on the large body of work characterizing differences within epiphytic microhabitats.

@@

We have demonstrated that growth habit, and several traits associated with it, show significant phylogenetic signal in ferns. Furthermore, we find that epiphytes are constrained by size while being remarkably diverse in other traits. Finally, we observed drier conditions in the canopy relative to terrestrial habitats. Taken together, our results support a scenario of environmental filtering acting to limit size and phylogenetic diversity in epiphytic ferns, while simulataneously enabling diversification along other trait axes. This is consistent with a much more broadly sampled study finding that epiphytic and terrestrial ferns do not differ in rates of body size evolution [@Testo2018a]. The generality of this pattern could be further tested by investigating the functional traits of other groups with large epiphytic radiations such as orchids and bromeliads.

Recent studies [e.g., @Sundue2015; @Testo2016a; @Lehtonen2017; @Testo2018a] have called into question the scenario of rapid Cretaceous radiations of epiphytic ferns posited by @Schuettpelz2009 by failing to detect increased rates of diversification associated with epiphytic growth. Epiphytism itself is likely a complex trait requiring the evolution of multiple associated traits, and the timing and order of the evolution of such traits may be related to the complex diversification dynamics observed in these studies. Small size seems to be a prerequisite for epiphytic growth, as indicated both by our study and @Testo2018a. Here, we identify several traits in addition to plant size (\eg, pinna number, frond dissection, gametophyte morphotype) that may also belong to the suite of epiphytic traits. We hope future studies will build on our results by analyzing traits involved in epiphytic growth other than body size across densely sampled phylogenies to dissect how each may relate to diversification and growth habit. Additionally, future studies should focus on determining in more detail how niche space is partitioned within epiphytic vs. terrestrial habitats, which is clearly a major dividing line for ferns.

>>>>

Despite the ecological opportunities afforded by life in the canopy, including escape from disturbance [@Page2002] and abundant light for photosynthesis [@Benzing2000], epiphytes must contend with greater stress in the form of increased evaporation and extreme oscillations in temperature relative to terrestrial plants [@Zotz2001; @Lowman2012]. Associated with this change, epiphytic ferns have evolved various morphological adaptations that minimize water loss from their leaves, including shorter stipes [@WatkinsJr.2010], thicker leaves [@Kluge2007; @WatkinsJr.2007b], and protection with scales or hairs [@WatkinsJr.2006b]. Some lineages have specialized adaptations to epiphytic growth, such as humus-collection (e.g., *Asplenium*; @Benzing1990, *Drynaria*; @Janssen2005), vegetative desiccation tolerance (DT) (e.g., Hymenophyllaceae; @Shreve1911, *Pleopeltis*; @Pessin1925), CAM photosynthesis (e.g., *Pyrrosia*; @Wong1976), and symbiotic relationships with ants (e.g., *Antrophyum*; @WatkinsJr.2008, *Lecanopteris*; @Gay1993a). Many of these same adaptations are found in epiphytic seed plants [@Benzing1990]. However, the evolution of epiphytism in ferns was more complex given that ferns rely on a free-living gametophyte [@Tsusumi2013]. The fern gametophyte is anatomically simple, and differs from the sporophyte in anatomy, morphology, and physiology. Such differences create a system of 'dual ecology' where selective pressures acting on one stage are different than those acting on another. For example, DT is prevalent in gametophytes of multiple lineages of epiphytic ferns, but is rare in sporophytes [@WatkinsJr.2007c]. Despite evidence for adaptations to epiphytism in multiple lineages in both sporophytes and gametophytes, no studies to our knowledge have analyzed the evolution of such traits across the fern phylogeny including both life stages and representatives of all five Cretaceous epiphytic radiations *sensu* [@Schuettpelz2009].

@@

We calculated community-weighted means (CWMs) and standard deviations of quantitative traits by using the overall mean trait value of each species weighted by their abundances in each community.

We used distance-based methods to analyze community phylogenetic and functional diversity. Mean phylogenetic distance (MPD) and mean nearest taxon distance (MNTD) describe the mean distance between all species in a community (MPD) or the mean distance between the closest pairs of species in a community (MNTD) [@Webb2000; Webb2002]. Although these metrics were originally formulated for phylogenetic distances, they can be applied in the same way to trait-based distances, enabling a direct comparison between the two [@Sessa2018]. For phylogenetic distances, we used the tree of @Nitta2017. For functional distances, we log-transformed and scaled traits as in the PCA, excluded frond width and length (both frond size traits highly correlated with stipe length), then calculated the distance matrix using Gower's dissimilarity coeffecient [@Gower1971] with the \function{gowdis} function in the R pacakge \package{FD} [@Laliberte2011], with equal weighting for sporophyte and gametophyte traits. To test for the effect of different trait selection schemes, we compared the results of running the same analyses for various sets of traits with and without log-transformation (Supporting Information).

Raw values of MPD and MNTD are statistically related to species richness [@Swenson2014]; to obtain metrics that are uncorrelated with species richness, we calculated their standard effect sizes (SES) by comparing observed values to a distribution of \num{999} randomized communities. We constructed null communities using the "independent swap" algorithm [@Gotelli2000]. We included all species of ferns from Moorea and the neighboring island of Tahiti with genetic data available (`r Ntip(phy)` \si{\species} total; [@Nitta2017]) for phylogenetic null communities, and all species with trait data available (`r fern_traits %>%pull(species) %>% n_distinct`) for functional null communities. We used the \function{ses.mpd} and \function{ses.mntd} functions in the R package \package{picante} [@Kembel2010] to calculate phylogenetic and functional diversity metrics. Hereafter, "MPD~phy~" and "MNTD~phy~", and "MPD~func~" and "MNTD~func~", refer to the standard effect sizes of MPD and MNTD as measured with phylogenetic and functional trait distances, respectively.

We tested the effects of climate and growth habit on diversity metrics using full-subsets multiple regression with general additive models (GAMs) as implemented in the \package{FSSgamm} R package [@Fisher2018]. For each diversity metric, we built a set of models (Gaussian family, with site as a random effect) taking into account all combinations of explanatory variables and their pairwise interactions, and ranked these by AIC~c~ to find the models that best explain the data. Our dataset included `r nrow(div_metrics_all) - nrow(climate_select)` communities that lacked climate data and could not be included in full-subsets multiple regression analysis. Therefore, we also constructed linear models for each diversity metric related to elevation including all sites using the same model selection procedure as in the climate analysis. We checked for spatial autocorrelation in the model residuals with Moran's *I* [@Moran1950] as implemented in the \package{spdep} R package [@Bivand2013].

@@

Results for functional community diversity contradicted our hypotheses; epiphytic communities were more functionally diverse than terrestrial as measured by both MPD and MNTD (\fig \@ref(fig:comm-div)). However, this result is dependent on the traits used to measure functional diversity; when the same analyses were done using only size-related traits (scaled, untransformed stipe length, frond length and width, and rhizome diameter), the opposite result was observed (\fig S1). Futhermore, terrestrial species show a much wider range of values for size traits (\fig \@ref(fig:cwm-div)), and neither functional group appears to be restricted to a smaller area in trait space (\fig \@ref(fig:pca)). This suggests that epiphytic ferns are subject to environmental filtering for size-related traits, while simultaneously diversifying along other trait axes. This could be maintained by niche partitioning within epiphytic habitats, which may be more diverse than terrestrial for a given area [@Rojas-Hernandez2018]. Epiphytic habitats are highly heterogeneous, and microclimatic conditions vary with height along the host and distance from the trunk [@Hietz1995]. The vertical stratification of epiphytic species has been well documented in ferns [\eg, @Parra2009; @WatkinsJr.2006a; @Acebey2017] and broader studies [\eg, @Kromer2007; @Zotz2007]. The goal of our study was to quantify differences between epiphytic and terrestrial fern communities; future research should focus on bringing comparative phylogenetic and community ecological methods to bear on the large body of work characterizing differences within epiphytic microhabitats.

@@

We have demonstrated that growth habit, and several traits associated with it, show significant phylogenetic signal in ferns. Furthermore, we find that epiphytes are constrained by size while being remarkably diverse in other traits. Finally, we observed drier conditions in the canopy relative to terrestrial habitats. Taken together, our results support a scenario of environmental filtering acting to limit size and phylogenetic diversity in epiphytic ferns, while simulataneously enabling diversification along other trait axes. The generality of this pattern could be further tested by investigating the functional traits of other groups with large epiphytic radiations such as orchids and bromeliads.

Recent studies [e.g., @Sundue2015; @Testo2016a; @Lehtonen2017; @Testo2018a] have called into question the scenario of rapid Cretaceous radiations of epiphytic ferns posited by @Schuettpelz2009 by failing to detect increased rates of diversification associated with epiphytic growth. Epiphytism itself is likely a complex trait requiring the evolution of multiple associated traits, and the timing and order of the evolution of such traits may be related to the complex diversification dynamics observed in these studies. Small size seems to be a prerequisite for epiphytic growth, as indicated both by our results and the global study of @Testo2018a. Here, we identify several traits in addition to plant size (\eg, pinna number, frond dissection, gametophyte morphotype) that may also belong to the suite of epiphytic traits. We hope future studies will build on our results by analyzing traits involved in epiphytic growth other than body size across densely sampled phylogenies to dissect how each may relate to diversification and growth habit. Additionally, future studies should focus on determining in more detail how niche space is partitioned within epiphytic vs. terrestrial habitats, which is clearly a major dividing line for ferns.
